---
title: "Ghana_YF_ENM"
output: html_document
date: "2024-05-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following R markdown file contains the R scripts used in the supplementary ENM analysis for the study: Yellow Fever in Ghana: Predicting Emergence and Ecology from Historical Outbreaks by Judson et al. 2024

The framework for tuning and evaluating the models was informed by the ENMEval R package and ENMEval 2.0 vignette by Kass et al. 2023 (https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html)

```{r}
library(ENMeval)
library(raster)
library(dplyr)
```

```{r}
library(sf)
library(ggplot2)
```

```{r}
library(ecospat)
```

```{r}
set.seed(15) #set random seed in order to reproduce this analysis
```

Load covariates and occurrence data
```{r}
folder_path <- "variables_asc" 
envs.files <- list.files(path = folder_path, pattern = "*.asc", full.names = TRUE)
envs <- raster::stack(envs.files) #load raster stack of covariates
```

```{r}
occs_all <- read.csv("occ_all.csv") #read in occurrences (all, urban, and savanna)
occs_sv <- read.csv("occ_savanna.csv")
occs_u <- read.csv("occ_urban.csv")
```


Mapping covariates and occurrences for demonstration purposes
```{r}
occs.sf <- sf::st_as_sf(occs_all, coords = c("long","lat"), crs = raster::crs(envs)) #make occurrences an sf object
```

```{r}
plot(envs[[1]], main = names(envs)[1]) #plotting bio1 (annual temperature) as an example 
plot(occs.sf, add = TRUE) # adding occurrences
```

Create background sampling points
```{r}
# Randomly sample 10,000 background points from one background extent raster 
# (only one per cell without replacement). 
bg <- dismo::randomPoints(envs[[1]], n = 10000) %>% as.data.frame()
colnames(bg) <- colnames(occs_all)
```
```{r}

plot(envs[[1]])
points(bg, pch = 20, cex = 0.2) #plotting background points
```

#Model containing all YF occurrences
```{r}
m_all <- ENMevaluate(occs = occs_all, envs = envs, bg = bg, 
                      algorithm = 'maxnet', partitions = "jackknife", 
                      tune.args = list(fc = c("L", "Q", "LQ"), rm = 1:5)) 
#K-1 leave-one-out cross validation method
#feature classes (fc) : linear, quadratic, linear * quadratic (interaction)
#regularization multipliers (rm): 1-5  (see Merow 2013 for discussion on fc and rm settings)
#note the pop density raster has additional cells that are NA in the other layers, so these are converted to NA in the pop density raster
```
```{r}
m_all
```
```{r}
# Overall results
m_all.res<-eval.results(m_all)
m_all.res
```

```{r}
#select models with the lowest average test omission rate, and to break ties, with the highest average validation AUC (Radosavljevic & Anderson 2014, Kass et al. 2020)
opt.seq_all <- m_all.res %>% 
  filter(or.10p.avg == min(or.10p.avg)) %>% 
  filter(auc.val.avg == max(auc.val.avg))
opt.seq_all #optimal model parameters fc.Q_rm.2 (auc.val.avg=0.73, or.10p.avg=0.13, delta.AICc =1.18)
#delta.AICc of <2 indicates statistically similar to others with the lowest AICc
```
```{r}
#select model with lowest AICc, delta.AICc is the difference between the AICc of a given model and the AICc of the model with the lowest AICc
opt.aicc_all <- m_all.res %>% filter(delta.AICc == 0)
opt.aicc_all 
```

```{r} 
mod.seq_all <- eval.models(m_all)[[opt.seq_all$tune.args]] #select optimal model

mod.seq_all$betas # non-zero coefficients in model
plot(mod.seq_all, type = "cloglog") #plot response curves
```

```{r}
pred.seq_all <- eval.predictions(m_all)[[opt.seq_all$tune.args]] #prediction based on selected optimal model above
plot(pred.seq_all)
points(eval.occs(m_all), pch = 21, bg = eval.occs.grp(m_all))
```

```{r}
writeRaster(pred.seq_all, filename = "YF_all.tif", format = "GTiff", overwrite=TRUE) #save prediction as raster TIFF
```


#Savanna models
```{r}
m_sv <- ENMevaluate(occs = occs_sv, envs = envs, bg = bg, 
                      algorithm = 'maxnet', partitions = "jackknife", 
                      tune.args = list(fc = c("L", "Q", "LQ"), rm = 1:5))
```

```{r}
m_sv
```
```{r}
# Overall results
m_sv.res<-eval.results(m_sv)
m_sv.res
```

```{r}
#select model with lowest AICc, delta.AICc is the difference between the AICc of a given model and the AICc of the model with the lowest AICc
opt.aicc_sv <- m_sv.res %>% filter(delta.AICc == 0)
opt.aicc_sv
```

```{r}
#select models with the lowest average test omission rate, and to break ties, with the highest average validation AUC (Radosavljevic & Anderson 2014, Kass et al. 2020)
opt.seq_sv <- m_sv.res %>% 
  filter(or.10p.avg == min(or.10p.avg)) %>% 
  filter(auc.val.avg == max(auc.val.avg))
opt.seq_sv

```
```{r} 
comp.seq_sv <- m_sv.res%>%
  filter(tune.args=='fc.Q_rm.2')
comp.seq_sv #select model with same feature classes and regularization multipliers as the model with all occurrences 
```

```{r}
mod.seq_sv <- eval.models(m_sv)[[opt.seq_sv$tune.args]] #using optimal model based on omission rate 

mod.seq_sv$betas ## non-zero coefficients in model
plot(mod.seq_sv, type = "cloglog")
```

```{r}
pred.seq_sv <- eval.predictions(m_sv)[[opt.seq_sv$tune.args]]
plot(pred.seq_sv)
points(eval.occs(m_sv), pch = 21, bg = eval.occs.grp(m_sv))
```

```{r}
writeRaster(pred.seq_sv, filename = "YF_sv.tif", format = "GTiff", overwrite=TRUE)
```

